Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-11-23 paper/manuscript.tex
Changes to uncertainty quantification section and suggestions for restructuring by @mrshirts

Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 
While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.

@@

\begin{figure*}
    \centering\includegraphics[width=0.85\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods and a typical workflow for executing a relative binding free energy calculation (yellow box).} 
    \textbf{A}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{B}: Change in free energy due to a conformational change of the molecule across a high barrier. \textbf{C}: Absolute free energies of binding of a small molecule to a host (e.g. protein). \textbf{D}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{E}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol. \textbf{F}: Effect of mutations of protein or host residues on free energies of binding. 
    \textbf{G.1}: the protein structure of the system is \textit{prepared} such that all erroneous patterns are resolved and proper protonation is applied. All query ligands are positioned into the binding site. \textbf{G.2}: $N=6$ compounds are predicted in this example. \textbf{G.3}: $n=7$ transformations (edges) are chosen in this example. \textbf{G.4}: for each transformation, a $\lambda$ decoupling parameter is applied to guide the transformation. \textbf{G.5}: for each transformation and $\lambda$ window a simulation is run. \textbf{G.6}: for each transformation, the relative binding free energy ($\Delta\Delta$G$_{bind}$) is calculated given $\lambda$. \textbf{G.7}: Each transformation (edge) has a corresponding binding free energy estimation. \textbf{G.8}: a final calculation is done to estimate the absolute binding free energy of each compound using for example a maximum-likelihood estimator. \textbf{G.9}: Absolute binding free energy estimates are ready to be used for prioritization in synthesis. \label{fig:fig_what_is_alchemy}
    }
\end{figure*}

@@

This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

 We assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
 Furthermore, basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association are essential. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead this guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I select an appropriate alchemical protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

@@

In this guide, some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations can be used to generate an initial small molecule binding pose to start alchemical simulations. This will require some basic familiarity on how to perform docking to generate reasonable simulation starting points~\cite{grinter2014challenges}.

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative). We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}. Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.

@@

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
% Add something on RBFE calculations across engines.
% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
% - other references?

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative, bhati2017, bieniek2021} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
%TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.

@@

The NEQ approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations. Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides a simple way to combine different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations. 
Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}. Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications. They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.

@@

%Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future. 
%Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.

@@

All of the estimators above have uncertainty estimates associated with them.  In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states. For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower. However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected. These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves. However, when the underlying data is noisy, this approximation is not very good.

@@

\paragraph{Recommendations}
\begin{itemize}
\item We recommend MBAR if all energy differences are available. It is the lowest variance unbiased free energy estimate given samples from multiple states.
\item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates. Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e. between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$. Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark} 
\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one is not careful.~\cite{paliwal2011benchmark}
\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used. If careful, it is not necessarily much worse than the other methods, but it always introduces some degree of binning error.
\item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
\end{itemize}

@@

To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Fig.~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
\begin{eqnarray}
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
\end{eqnarray}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice. A very similar approach involves minimising the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.

@@

Note that the use of the effective reduced potential is not guaranteed to pick up on all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.

\paragraph{Cautions in automating equilibration detection}
For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of configurational space, and may itself lead to biased estimates. 
This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.

@@

If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state. If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state. Conveniently, maximising the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.

@@

Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an Ito integral with Iso isomoetry for error propogation.

@@

>>>>

Alchemical free energy calculations are a useful tool for predicting free energy differences associated with the transfer of molecules from one environment to another.
The hallmark of these methods is the use of ``bridging'' potential energy functions representing \emph{alchemical} intermediate states that cannot exist as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 
While these methods are highly flexible, care must be taken in avoiding common pitfalls to ensure that computed free energy differences can be robust and reproducible for the chosen force field, and that appropriate corrections are included to permit direct comparison with experimental data.

@@

\begin{figure*}
    \centering\includegraphics[width=0.85\linewidth]{figures/fig1_what_is_alchemy/Figure.pdf}   
    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods and a typical workflow for executing a relative binding free energy calculation (yellow box).} 
    \textbf{A}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{B}: Change in free energy due to a conformational change of the molecule across a high barrier, such as the torsion reaching from the aromatic ring to the hydroxyl group. \textbf{C}: Absolute free energies of binding of a small molecule to a host (e.g. protein). \textbf{D}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{E}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol. \textbf{F}: Effect of mutations of protein or host residues on free energies of binding. 
    \textbf{G.1}: the protein structure of the system is \textit{prepared} such that all erroneous patterns are resolved and proper protonation is applied. All query ligands are positioned into the binding site. \textbf{G.2}: $N=6$ compounds are predicted in this example. \textbf{G.3}: $n=7$ transformations (edges) are chosen in this example. \textbf{G.4}: for each transformation, a $\lambda$ decoupling parameter is applied to guide the transformation. \textbf{G.5}: for each transformation and $\lambda$ window a simulation is run. \textbf{G.6}: for each transformation, the relative binding free energy ($\Delta\Delta$G$_{bind}$) is calculated given $\lambda$. \textbf{G.7}: Each transformation (edge) has a corresponding binding free energy estimation. \textbf{G.8}: a final calculation is done to estimate the absolute binding free energy of each compound using for example a maximum-likelihood estimator. \textbf{G.9}: Absolute binding free energy estimates are ready to be used for prioritization in synthesis. \label{fig:fig_what_is_alchemy}
    }
\end{figure*}

@@

This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this purpose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations can take advantage of currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 This guide can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers including alchemical free energy calculations.

In this best practices document, we assume that novice practitioners have at least moderate experience with molecular simulation concepts and use of simulation packages. 
We also assume basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association. If you feel unfamiliar with some of these concepts, good starting points can be found in these references~\cite{braun2019best, grossfield2018best, klimovich2015guidelines, shirts2012best}.

 While reading this Best Practices guide, it is important to bear in mind \emph{this is not a review} of all free energy calculation methods at the cutting edge of current research.
Instead this guide aims to answer the following questions:
\begin{itemize}
    \item Is my problem suitable for an alchemical calculation? 
     \item How do I select an appropriate simulation protocol? 
     \item What software tools are available to perform alchemical calculations? 
     \item How should I analyze my data and report uncertainties? 
\end{itemize}

@@

In this guide, some other background information may be needed depending on the nature of the alchemical project. For example, often, if binding poses are not known, docking calculations or co-folding models can be used to generate an initial small molecule binding pose to start alchemical simulations. This will require some basic familiarity on how to perform docking or using generative models. to generate reasonable simulation starting points~\cite{grinter2014challenges}.

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation (Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
We will give particular focus to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative). We primarily focus on free energy calculations using simulations performed at \emph{equilibrium} in this Best Practices guide, as best practices for these are more developed, and non-equilibrium techniques may warrant their own guide as such practices evolve.

@@

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.

Simulations carried out with molecular dynamics or Monte Carlo techniques are fundamentally stochastic calculations, and two simulations can never match exactly.  Even though Newton's equations of motion are deterministic, molecular systems are chaotic, and very small changes, like differences in rounding of floating point numbers due to different software versions or architectures, let alone different random number seeds used to initialize the velocities will lead to different results obtained with each repetition of a simulation. Any sort of agreement will only be up to this stochastic source of uncertainty, and one needs to have an understanding of how accurate one needs comparison to be, as well as a proper handling of statistics to decide if differences in simulations are statistically significant or not.  Because free energy differences are calculated from ensemble averages, collecting more data will result in improved agreement in almost all cases, thanks to the central limit theorem, but the amount of simulation needed to converge to a desired level of statistical precision varies dramatically on the system. So the key consideration is how reproducible alchemical free energy calculations are in practice, and how large a difference between two measurements is required to be scientifically useful.

There are two types of reproducibility that are important to examine.  One is reproducibility between different research groups attempting to reproduce the same results, which is relevant when new users might be trying to see how well their calculations match with published calculations. The second is reproducibility between one researcher running the same calculation, up to differences in starting configurations, multiple times.  This can be particularly relevant in determining the statistical uncertainty of the calculation, as discussed in Sec~\ref{subsec:uncertainty}.

%Even within this restriction, finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision.

The differences in free energies between different research groups, due to differences in choices of simulation parameters, simulation packages, and analysis treatments can be significant. Issues of reproducibility of free energy calculations across different simulation packages have recently attracted attention ~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022, herz2025modular}. Greater variability is expected between simulations carried out between packages, as opposed to ones carried out with the same package, due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so.  For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.   For benchmark simulations of some relative binding free energies, differences of {\color{red}Can we put some more citations here?}

A different situation is the amount of reproducibility researchers should expect when performing repeated simulations of their own systems, perhaps with the same starting point. In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol possible~\cite{rizzi2019sampl6}, and agreement under 0.05 kcal/mol routine for simulations lasting 10's of ns~\cite{paliwal2011benchmark} {\color{red}This could use some more citations here from other studies.} However, for complex systems, the reproducibility can be signficantly worse, as two simulation of the same system can explore very different configurational ensembles.  If the timescales of the slow scales of motion are comparable to or longer than the timescales of the simulation, then differences of up to multiple kcal/mol can be possible for absolute and even relative free energy calculations.

In fact, the amount of reproducibility between repeated calculations is an important measure of whether a particular choices of simulation methods and approaches are appropriate.  Large differences between results carried out with different initial velocities, or better yet started from points along a single initial long simulations can indicate that the sampling of the system is particularly hard and will require additional simulation and/or advances sampling methods.  In fact, journals such as the Journal of Chemical Information and Modeling require that any results be confirmed by simulations run in at least triplicate. The question of how to combine results from ensembles of simulations for lowest bias and best reporting of statistical uncertainty are discussed in section~\ref{subsec:uncertainty}.

%MRS: some good points here below, but this isn't the right place, since this is explicitly talking about reproducibility between simulations, i.e. if you run a simulation, how different do you expect it to be. This MOTIVATES a discusion of ensemble averaging in some other section for robustness, but should be  separate sections discussing ensembles vs. simgle trajectories, putting all these differences in the same place, rather than sprinkled throughout.  Also, some of the claims are too broad as stated, and it included information about free energy calculation methods that doesn't fit here.

%MD trajectories are extremely sensitive to their starting conditions and hence single simulationscan never be reproducible~\cite{coveney2016, MDbook2025, caves_locally_1998}. Thus, it is recommended to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods, where the ensemble size can be adjusted based on available computational resource and required precision~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}.

%MRS: Should be in a separate section. 
%The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand.

@@

% Add something on RBFE calculations across engines.
% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
% - other references?

@@

\paragraph{Ring breaking and forming} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative, bhati2017, bieniek2021} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
%TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.

@@

The NEQ approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations. Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides a simple way to combine different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates.

%MRS: I think this below is a little too detailed for a general introduction of NEQ, as we should be talking about general strategies, not specific ones.  Probably we should be putting in a few references for key NEQ uses so readers could follow up, however, THat would be useful to have here.

%MRS: was this sentence the one with the violated assumptions?  please check, I am note sure. 
%Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations.

%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}. Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications. They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.

@@

%Currently, we recommend using Hamiltonian replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme} \textbf{B}). If these are not available in the code of choice, running independent simulations at different $\vec{\lambda}$ values can be acceptable, especially when configurational sampling is fast (Fig.~\ref{fig:fig_sampling_scheme} \textbf{A}). Single replica schemes and non-equilibrium schemes are not as established yet because of potential failure modes, but are very promising for use in the near future. 
%MRS: doesn't below here. 
%Further, ensemble simulations are recommended irrespective of the choice of sampling scheme to ensure reliability of results~\cite{bhati2018, bhati2019, wan2020fep+, wade2022, wan2021uq, bhati2025, wan2023eqvsneq}.

%MRS: This should be broken up into two sections - 
% The second of which is "what information should be saved - that should really be separated. 
% The first one should be perhaps, " How much simulation data should I collect" and THIS is where we should put discussions of ensembles versus long siulations.  I think the question "how long should I run?" is not the right questions, because as Agastya points out, how to break them up between MULTIPLE runs is an important discussion we should have.

@@

All of the estimators above have uncertainty estimates associated with them.  In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states. For example, for a series of hydration free energy calculations of model systems of various sizes analytical estimates of the error, bootstrap estimates, and standard deviations over 100 independent simulations started at different locations along a long initial simulations all agreed within a 5---10\%.~\cite{paliwal2011benchmark} For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower.

However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected.~\cite{???} These estimators of uncertainty are all derived from propagation of error formulae from the estimators themselves. When the underlying data is noisy, this approximation is not very good.

@@

\paragraph{Recommendations}
\begin{itemize}
\item We recommend MBAR if all energy differences are available. It is the lowest variance unbiased free energy estimate given samples from multiple states.
\item BAR is essentially just as good as MBAR for highly optimized $\vec{\lambda}$ intermediates. Specifically, if the $\vec{\lambda}$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e. between $i$ and $i+1$ and between $i$ and $i-1$), they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$. Thus MBAR does not actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal2011benchmark} 
\item TI usually gives similar values as MBAR implemented with sufficient numbers of intermediates, but quadrature errors that are hard to estimate beforehand  can occur if one does not carefully select $\lambda$ values such that the overlap between states is sufficiently high.~\cite{paliwal2011benchmark}
\item WHAM is an approximation to MBAR, and there are no compelling reasons it should be used. If bin widths are careful chosen and there are sufficient samples, it can given similar quality results to have the other methods, but it always introduces some degree of binning error, and determining when errors might occur can be challenging. 
\item Other variants, especially ones that adaptively determine the free energies can be useful in certain circumstances but beyond the scope of a Best Practices article.
\end{itemize}

@@

To minimize this effect, an initial portion of the simulation that is determined to be  \emph{equilibration}~\cite{braun2019best} should be omitted, with the idea of removing the most heavily biased initial portion of simulation data, but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step, rather than attempting to figure out beforehand how long equilibration will take.

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Fig.~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
\begin{eqnarray}
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
\end{eqnarray}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice. A very similar approach involves minimizing the standard error of the production region~\cite{whiteEffectiveTruncationHeuristic1997, sprattHeuristicsStartupProblem1998, hoadAutomatingWarmupLength2010, clarkRobustAutomatedTruncation2025}.

@@

Note that the use of the effective reduced potential is not guaranteed to detect all slow relaxation processes that may be coupled to the alchemical free energy, but the simplicity of its computation means it is generally appropriate for most cases.

\paragraph{Cautions in automating equilibration detection}
Procedures like those described above can fail to remove bias if the simulations are simply not long enough to contain a large number of samples from true equilibrium either because they are very short, or because one is simulating slow processes. In such cases, this approach often simply selects a small final portion of the simulation, which may be contained in a single substate of configurational space, which may itself lead to biased estimates. 
This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that under well-sampled situations $T \gtrsim 20 t_0$.
If identifying an equilibration time that is only a fraction of the total time is not possible, advanced analysis techniques that assume only local equilibrium within kinetically separated metastable states, rather than global equilibrium over the entire phase space, such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper.

@@

If independent simulations at each $\lambda$ are used, the alchemical state $\vec{\lambda}$ may have a significant impact on the correlation time, and these simulations should be subsampled independently using a separate estimate of the statistical inefficiency $g$ for each alchemical state. If coupled simulations are used (such as a Hamiltonian replica exchange simulation), the replicas should undergo equivalent random walks in alchemical space, and the replicas can be subsampled with the same $g$ to generate an equal number of uncorrelated samples at each alchemical state. Conveniently, maximizing the effective sample size for automated equilibration detection produces an appropriate estimate of $g$ over the production region for automating this process.

@@

Beyond thermodynamic integration, the TIES method\cite{bhati2017} uses an It\^{o} integral with Iso isomoetry for error propogation.

@@

% notthing here?  It would actually be really nice to summarize this.
